Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-02 doc/types-pldi2013/conclusion.tex
minor edit
updates

Modern architectures have dramatically increased in complexity in recent years.  To program this class
of machines, new programming systems will be required that are capable about reasoning about the structure
of data including aliasing.  We have presented a type system for the Legion programming system that
reasons about aliasing using a combination of static and dynamic analyses.  Furthermore, we've proven
the soundness of our type system enables a hierarchical, distributed scheduling algorithm capable
of scaling on large, distribute memory machines.  This result demonstrates that allowing programmers
to provide descriptions of data that include aliasing is not prohibitive to high performance.

>>>>

%Modern architectures have dramatically increased in complexity in recent years.  To program this class
%of machines, new programming systems will be required that are capable about reasoning about the structure
%of data including aliasing.  
We have presented static and dymamic semantics for the Legion programming system,
showing how a combination of static and dynamic checks can be used to soundly and efficiently schedule
Legion programs in parallel, even in the presence of dynamically partitioned and aliased data.  We have
also given a novel compositional parallel semantics, permiting a precise treatment of relaxed coherence modes;
in particular we have shown the Legion design is sound even with relaxed coherence.

### 2012-11-02 doc/types-pldi2013/evaluation.tex
updates

We evaluate the Legion type system on four criteria:
\begin{itemize}
\item Expressivity: Can real applications be written (Section~\ref{subsec:expressivity})?
\item Overhead: Does it reduce dynamic checking costs (Section~\ref{subsec:overhead})?
\item Scalability: Can it enable hierarchical scheduling (Section~\ref{subsec:scalability})?
\item Performance: Does the performance increase from relaxed coherence modes 
warrant the increased complexity in the operational semantics (Section~\ref{subsec:relaxedperf})?
\end{itemize}
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model\cite{Legion12}.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

@@

\subsubsection{Fluid Simulation}
\label{subsec:fluid}
Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move within a regular grid of cells.  To perform operations in 
parallel the array of cells is partitioned.  Unlike
Circuit, Fluid creates and partitions regions before
allocating cells in them (i.e., partition-then-allocate).  Another difference is that Fluid
maintains separate regions for ghost cells rather using multiple partitions of
the regions containing shared data.  Region relationships are used for remembering
which regions are required for each grid.

\subsubsection{Adaptive Mesh Refinement}
\label{subsec:amr}
The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with sub-refinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every level of refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

%These applications directly demonstrate the need to for multiple, dynamic partitions
%of program data.
%These applications demonstrate Legion is capable of 
%expressing real applications with both regular (Fluid, AMR) and 
%irregular (Circuit) pointer data structures.
Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence we have
described are needed in different applications.  Region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multi-colorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

>>>>

We evaluate the design of Legion's static and dynamic semantics on four criteria:
expressivity (can real applications be written---Section~\ref{subsec:expressivity}),
overhead  (what are the dynamic checking costs---Section~\ref{subsec:overhead}),
scalability (can it enable hierarchical scheduling---Section~\ref{subsec:scalability}),
and performance  (does the performance increase from relaxed coherence modes 
warrant the increased semantic complexity---Section~\ref{subsec:relaxedperf}).
%Our evaluation uses a prototype implementation of the Legion language and programming model.
Our prototype implementation has two components: a type checker
for the language introduced in Section~\ref{sec:legioncore} and a C++ runtime library
for executing programs written in the Legion programming model.  All experiments
are conducted on the Keeneland supercomputer\cite{Keeneland}.  Each node of the Keeneland
machine consists of two Xeon 5660 CPUs, three Tesla M2090 GPUs, and 24 GB of DRAM.  Nodes
are connected by a QDR Infiniband interconnect.

@@

%\subsubsection{Fluid Simulation}
%\label{subsec:fluid}
Fluid is based on the {\tt fluidanimate} benchmark from the PARSEC 
suite\cite{bienia11benchmarking}.  Fluid simulates the flow of an incompressible fluid
using particles that move within a regular grid of cells.  To perform operations in 
parallel the array of cells is partitioned.  Unlike
Circuit, Fluid creates and partitions regions before
allocating cells in them. 
%(i.e., partition-then-allocate).  
Another difference is that Fluid
maintains separate regions for ghost cells rather using multiple partitions of
the regions containing shared data.  Region relationships are used for remembering
which regions are required for each grid.

%\subsubsection{Adaptive Mesh Refinement}
%\label{subsec:amr}

The third application is a Legion port of an adaptive mesh refinement (AMR) benchmark 
from BoxLib \cite{BoxLib}.  The algorithm solves the two
dimensional heat diffusion equation on a grid of cells using three levels of refinement with sub-refinements
randomly placed on the surface.  
%For each time step in the application, cells at the boundary of
%a refinement are interpolated from cells at a coarser level of refinement, energy fluxes between
%cells are computed, energy is transferred, and cells at a coarser level of refinement are restricted
%to the values of refined cells.  
Every level of refinement uses a separate region, which is partitioned several ways to support multiple views of
the cells.  One partitioning separates cells into pieces that can be updated in
parallel.  Additional partitions are created for viewing data from coarser and finer levels of
the simulation.  Two types of region relationships are created: one for describing the pieces at 
each level of refinement, and another for describing the relationship between pieces at different
levels of refinement.  
%These region relationships capture both intra- and inter-level disjointness information.  
The dynamic nature of AMR requires that regions be created and partitioned at runtime.  
%The many ways in which cells are used also requires multiple simultaneous
%partitions of regions.

%These applications directly demonstrate the need to for multiple, dynamic partitions
%of program data.
%These applications demonstrate Legion is capable of 
%expressing real applications with both regular (Fluid, AMR) and 
%irregular (Circuit) pointer data structures.

Being able to dynamically create and partition regions at runtime is crucial to Legion's
ability to handle applications that make runtime decisions about data organization (AMR).  Legion
is able to capture both allocate-then-partition (Circuit) and partition-then-allocate (Fluid) algorithms. 
Having multiple partitions of regions is necessary for describing the many ways 
that data can be accessed (Circuit, AMR).  All the types of privileges and coherence 
are needed in some application;  region relationships
are used in all applications.   Finally, all applications introduce aliasing of
data through both multi-colorings or having multiple partitions.  Our implementations of
these applications both type check and execute, proving that our type system is
sufficiently expressive to handle real-world applications.
% written in the Legion programming model.
%, region relationships convey disjointness information.
% despite
%regions being a first-class feature that cannot be fully tracked statically.

### 2012-11-02 doc/types-pldi2013/langdef.tex
updates

The enforcement of a (locally) valid interleaving at every step in the operational semantics
captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
a scheduling decision is quadratic in the number of function calls (more precisely, the number of 
privileges for those calls), it should be clear that a hierarchical scheduling algorithm will
be vastly more efficient than a global algorithm that must consider the privileges of every
concurrently executing expression.
However, a scheduling algorithm that makes decisions hierarchically must do so with limited
information, and must respect all required scheduling dependencies while ideally imposing as
few extra dependencies as possible.  The following theorem forms the basis of the 
Legion scheduler.

\texcomment{
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
permissible.  We do so by proving that any memory traces that can result from the evaluation
of the subexpression must be non-interfering (allowing parallel execution by
Lemma~\ref{lem:noninteffects}).
}
\begin{thm}
\label{thm:hiersched}
\rm
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$), must be
non-interfering.
\end{thm}
\texcomment{
\begin{proof}
Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
interleavings, each of which must include all the memory operations of each constituent trace.
By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
\end{proof}
} The Legion task scheduler uses Theorem~\ref{thm:hiersched} as follows: 
sibling function calls (those invoked within the same
function body) $e_1$ and $e_2$ are checked for non-interference of their (dynamic)
privileges.  Since $e_1$ and $e_2$ are guaranteed to be created on the node on which their parent
is running, no communication is required to perform the non-interference
test.  If they interfere they are executed in program order or serialized
depending on their coherence specifications; otherwise they are
considered for execution as parallel subtasks.    If $e_1$ and $e_2$ are 
determined to be non-interfering and are scheduled in parallel on different 
remote processors then Theorem~\ref{thm:hiersched}
guarantees that there is no communication required between $e_1$ and $e_2$
to perform non-interference tests between their sub-tasks.  Therefore,
the runtime requires no communication for scheduling.

Although determining that $e_1$
and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
non-interfering, the converse is not true.  It is possible for $e_1$
and $e_2$ to be judged to be interfering and be executed sequentially 
even when the subexpressions are non-interfering.  This loss of
precision can impact performance, and did in a few instances in our
experiments.  
%When the performance impact is significant, two options
%exist to improve matters.  
In cases where $e_1$ and $e_2$ are in fact
non-interfering, region partitions can be further refined to make the
disjointness visible in the privileges.  Alternately, the code can be
refactored to lift the subexpressions up to the same level, allowing
the dynamic test based on the region mapping to infer non-interference
in more cases.  
%Although this can increase the overhead of scheduling
%decisions, judicious use of this technique in practice has not shown
%any measurable performance degradation.

>>>>

Because testing non-interference of tasks is a pairwise operation, scheduling $n$ tasks can require ${\cal O}(n^2)$
tests.  Thus, a scheduler that must globally consider all pairs of tasks will be impractical for large machines 
and large numbers of tasks.  The following theorem, however, shows that Legion programs enjoy a locality property that
limits the scope of needed non-interference tests.

%The enforcement of a (locally) valid interleaving at every step in the operational semantics
%captures the hierarchical nature of the scheduler used by the Legion runtime.  As the cost of
%a scheduling decision is quadratic in the number of function calls (more precisely, the number of 
%privileges for those calls), it should be clear that a hierarchical scheduling algorithm will
%be vastly more efficient than a global algorithm that must consider the privileges of every
%concurrently executing expression.
%However, a scheduling algorithm that makes decisions hierarchically must do so with limited
%information, and must respect all required scheduling dependencies while ideally imposing as
%few extra dependencies as possible.  The following theorem forms the basis of the 
%Legion scheduler.

\texcomment{
Let $e_1$ and $e_2$ be well-typed Legion expressions using privileges $\Phi_1$ and
$\Phi_2$.  Now let $e_3$ be some subexpression of $e_1$ and $e_4$ be some 
subexpression of $e_2$, with corresponding privileges $\Phi_3$ and $\Phi_4$.
By Theorem~\ref{thm:parallelexec}, if the runtime can verify $\Phi_1 \nonint[M] \Phi_2$, then 
expressions $e_1$ and $e_2$ may be evaluated in parallel.  The evaluation of the subexpressions
$e_3$ and $e_4$ will therefore be allowed to run in parallel, and we must show that this is
permissible.  We do so by proving that any memory traces that can result from the evaluation
of the subexpression must be non-interfering (allowing parallel execution by
Lemma~\ref{lem:noninteffects}).
}
\begin{thm}
\label{thm:hiersched}
\rm
Let $e_1$ and $e_2$ be well-typed expressions using privileges $\Phi_1$ and
$\Phi_2$ respectively, where $\Phi_1 \nonint[M] \Phi_2$.  Let $e_3$ be a subexpression of $e_1$ and $e_4$ be a 
subexpression of $e_2$.  Any memory traces $E_3$ of $e_3$ and $E_4$ of $e_4$ resulting from evaluation of
$e_1$ and $e_2$ (with the usual consistent $M$, $L$, $H$, and $S$) are 
non-interfering.
\end{thm}
\texcomment{
\begin{proof}
Let $E_1$ be the result of an evaluation of $e_1$, with $E_3$ being the memory trace from the
evaluation of $e_3$ within the expression tree.  $E_1$ will be formed from a tree of valid
interleavings, each of which must include all the memory operations of each constituent trace.
By induction over the number of intermediate subexpressions between $e_1$ and $e_3$, we
can show that any memory operation $\epsilon_3$ that is in $E_3$ must also be in $E_1$.
Similarly, any memory operation $\epsilon_4$ that is in $E_4$ must also be in $E_2$.  We have
$E_1 \nonint E_2$ from Lemma~\ref{lem:nonintpriv}, so and therefore $\epsilon_3 \nonint \epsilon_4$.  Since
this holds for all $\epsilon_3$ in $E_3$ and all $\epsilon_4$ in $E_4$, we have $E_3 \nonint E_4$.
\end{proof}
} The Legion task scheduler uses Theorem~\ref{thm:hiersched} as follows: 
sibling function calls (those invoked within the same
function body) $e_1$ and $e_2$ are checked for non-interference of their (dynamic)
privileges.  Since $e_1$ and $e_2$ are guaranteed to be created on the node on which their parent
is running, no communication is required to perform the non-interference
test.  If they interfere they are executed in program order or serialized
depending on their coherence specifications; otherwise they are
considered for execution as parallel subtasks.    If $e_1$ and $e_2$ are 
determined to be non-interfering and are scheduled in parallel on different 
remote processors then Theorem~\ref{thm:hiersched}
guarantees that there is no communication required between $e_1$ and $e_2$
to perform non-interference tests between their sub-tasks.  Therefore,
the runtime requires no communication for scheduling.

%
% FIXME
% This is a nice and useful discussion, but also not on the main track and good candidate for space savings.
% Marking this with a FIXME so we can reconsider if we like.
%
% 
%Although determining that $e_1$
%and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
%non-interfering, the converse is not true.  It is possible for $e_1$
%and $e_2$ to be judged to be interfering and be executed sequentially 
%even when the subexpressions are non-interfering.  This loss of
%precision can impact performance, and did in a few instances in our
%experiments.  
%When the performance impact is significant, two options
%exist to improve matters.  
%In cases where $e_1$ and $e_2$ are in fact
%non-interfering, region partitions can be further refined to make the
%disjointness visible in the privileges.  Alternately, the code can be
%refactored to lift the subexpressions up to the same level, allowing
%the dynamic test based on the region mapping to infer non-interference
%in more cases.  
%Although this can increase the overhead of scheduling
%decisions, judicious use of this technique in practice has not shown
%any measurable performance degradation.

### 2012-11-02 doc/types-pldi2013/related.tex
updates

Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges, 
%and we have reused some DPJ notation, 
there are differences stemming from DPJ's static approach.
Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 
dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
partition schemes must be statically decided.  Legion allows 
programmers to create multiple partitions of the same region to give different 
views onto the same data, which is not possible in DPJ.  DPJ supports both
exclusive and atomic tasks which are similar to Legion's coherence modes, but only
allows specification at the coarser granularity of tasks and not individual regions.
DPJ does not support any modes analogous to simultaneous coherence.
%DPJ only supports the 
%equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
%while Legion provides safe execution even in simultaneous environments.  
There is also a difference in emphasis: DPJ requires shared memory, while Legion 
is designed for distributed heterogeneous machines.

@@

There have been many type and effect systems for ownership types
\cite{Boyapati03} including ones that leverage nested regions for describing
relationships \cite{Clarke02,Cameron07}.  However, ownership type and effect systems
are primarily used for reasoning about determinism in object oriented languages and
don't capture the range of disjointness properties that can be specified in Legion.

Reasoning about disjoint heap data is the strong suit of separation logic \cite{Reynolds02}.  
Concurrent separation logic\cite{Brookes04} has been 
used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
a mechanism for reasoning about independence\cite{Hayman06}.
While we have borrowed some separation logic notation, we ultimately chose to use a 
privileges system as our primary formalism because separation logic does not easily
support reasoning about the interleaving of operations to aliased regions of memory.

>>>>

Deterministic Parallel Java (DPJ) is the only other region-based parallel system of which we are
aware\cite{Bocchino09}.  While there are similarities between DPJ's effects and Legion privileges, 
%and we have reused some DPJ notation, 
there are differences stemming from DPJ's static approach.
Regions in Legion are first-class and can be created, partitioned, packed, and unpacked 
dynamically, allowing programmers to compute data organization at runtime; like Sequoia, DPJ
partition schemes are static.  Legion allows 
programmers to create multiple partitions of the same region to give different 
views onto the same data, which is not possible in DPJ.  DPJ supports both
exclusive and atomic tasks which are similar to Legion's coherence modes, but only
allows specification at the coarser granularity of tasks and not individual regions.
%DPJ does not support any modes analogous to simultaneous coherence.
%DPJ only supports the 
%equivalent to Legion's exclusive and atomic coherence modes \cite{Bocchino11} 
%while Legion provides safe execution even in simultaneous environments.  
%There is also a difference in emphasis: DPJ requires shared memory, while Legion 
%is designed for distributed heterogeneous machines.

@@

There have been many type and effect systems for ownership types
\cite{Boyapati03} including ones that leverage nested regions for describing
relationships (e.g., \cite{Clarke02}).  However, ownership type and effect systems
are primarily used for reasoning about determinism in object oriented languages and
don't capture the range of disjointness properties that can be specified in Legion.
Reasoning about disjoint data is the strong suit of separation logic \cite{Reynolds02}.  
%Concurrent separation logic\cite{Brookes04} has been 
%used both to parallelize sequential programs\cite{Raza09,Gotsman07} and to provide 
%a mechanism for reasoning about independence\cite{Hayman06}.
While we have borrowed some separation logic notation, we ultimately chose to use a 
privileges system as our primary formalism because separation logic does not easily
support reasoning about the interleaving of operations to aliased regions of memory.
